### 2017-02-06 manuscript.rmd
made a complete pass at text, now compute BM model parameters once

@@

		# Estimate parameters
		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
		names( tau_original ) = phy$tip.label
		brownian_model = fitContinuous( phy, tau_original )

@@

There is now considerable interest in comparing functional genomics data across species, with the hope of providing an integrated understanding of genome and phenotype evolution. Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species. This has the potential to mislead conclusions about the evolution of genome function, and is a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of the phylogenetic trees themselves. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor. This violates the fundamental assumption of independence of observations in standard statistical methods. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to functional genomics. Most Comparative Functional Genomics (CFG) studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig. 1A). These do not account for the fact that evolutionary relationships are expected to explain much of the structure of variation across species. This leaves CFG studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods explicitly model trait change along the branches of evolutionary trees. The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

The second CFG study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases for these ten animals (Fig. XX). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we directly address the analyses themselves by examining the structure of the pairwise comparisons.

@@

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

Results in KMRR are based on pairwise comparisons (Figure 1A) between Tau within each gene family. They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs".

@@

We reanalyzed the KMRR study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [@Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

To test the ortholog conjecture it is essential to have a null hypothesis that makes predictions distinct from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events [@Studer:2009bm] (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

@@

When we reanalyze the data with phylogenetic independent contrasts, we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

Given the ortholog conjecture, we would expect to see larger contrasts following duplication events than speciation events. Instead, we find no such distribution (Figure XXKMRR). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our results therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolution following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length. We found no evidence that either affected our results (Supplementary Materials). We also examined the sensitivity of the results to the calibration times used for speciation events on the gene trees. This is important because it is expected that genes from separate species trace backward in time to a common ancestor at older times than the splitting times of the species to which they belong [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. In replicate analyses we added random noise to the calibration times and found that it does not impact our inability to reject the null hypothesis (Supplementary Materials).

@@

	pairwise_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic()

@@

> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base. Each circle indicates the Tau Pearson correlation (rho) for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs. We find the same here, with correlation coefficients of R=`r round( ortholog_r, 3)` between ortholog pairs and `r round( paralog_r, 3)` for paralog pairs.

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` My, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` My. As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly related paralogs arising from these older duplication events (Supplementary Figure XXAscertainment a) are the reason for the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the observed pattern is very different. The correlation of Tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog correlation, which has R=`r round( ortholog_r, 3) `.

@@

	# Replace the actual Tau values with simulated Tau values, using parameters 
	# fitted from each tree
	gene_trees_sim_null = lapply( gene_trees_calibrated, sim_tau )

@@

	# Replace the actual Tau values with simulated Tau values, using parameters 
	# fitted from each tree. 
	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )

@@

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Age" ) +
			ylab( "Rho" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null simulation, pairwise" ) +
			theme(legend.position="none")

@@

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Age" ) +
			ylab( "Rho" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC simulation, pairwise" ) +
			theme(legend.position="none")

@@

We simulated a data set under the null model that tau evolves at the same rate following duplication and speciation events. Under the null model, pairwise analysis results (Figure XXSim a) are very similar as for the observed data, with higher rho for orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`). In the phylogenetic analysis of the null simulation, contrasts show similar divergence following speciation and duplication (Figure XXSim b) and does not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`).

We next simulated a data set under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The results of this heterogeneous model are nearly indistinguishable from the null model  (Figure XXSim c), and also have a higher rho for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`). In the phylogenetic analysis of the ortholog conjecture simulation, contrasts XX (Figure XXSim c) and XX (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).

@@

We find that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and history of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].

There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi]. Our results are consistent with multiple studies that have not found support for the ortholog conjecture [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk]. Another study found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has important biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].

>>>>

	#' Estimate Tau evolution parameters from a tree::treedata tree
	#' and add them back into the tree object.
	#' 
	#' @param nhx A phylogenetic tree and associated Tau values 
	#' as a treeio::treedata object
	#' @param ... Additional arguments to pass to geiger::fitContinuous()
	#' @return A phylogenetic trees as a treeio::treedata object, 
	#' with estimated brownian motion model parameters stored 
	#' in @phylo$model
	#' @export
	add_model_parameters = function( nhx, ... ) {
		phy = nhx@phylo
		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
		names( tau_original ) = phy$tip.label
		nhx@phylo$model = fitContinuous( phy, tau_original, ... )

@@

		# Get parameter estimates
		brownian_model = nhx@phylo$model

@@

With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species. Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incoporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence of the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future funcitonal genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics. Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig. 1A). This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees. The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendancy of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

@@

The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig. XX). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we address the analyses themselves by examining the structure of the pairwise comparisons.

@@

The KMRR study considered several publicly available datasets of gene expression across tissues and species. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

The KMRR analyses are based on pairwise comparisons (Figure 1A) between Tau within each gene family. They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].

@@

We reanalyzed the KMRR study using phylogenetic comparative methods. We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du]. This dataset is the best sampled in their analyses. It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.

@@

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for  `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

It is essential to have a null hypothesis that makes a distinct prediction from the prediction of the hypothesis under consideration. A suitable null hypothesis in this case  is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm]. Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

We did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events. Such a bias might obscure support for the ortholog conjecture. We focused on two possible sources of bias, node depth and branch length. We found no evidence that either affected our results (Supplementary Materials). We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees. This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Materials).

@@

	pairwise_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" )

@@

> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base. Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1, they present a higher Tau correlation coefficient between ortholog pairs than paralog ortholog pairs. We find the same here, with correlation coefficients of `r round( ortholog_r, 3)` for orthologs and `r round( paralog_r, 3)` for paralogs.

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure XXAscertainment a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`. This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.

```{r estimate_tau_model, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
	# Estimate model parameters and add them to the tree objects
	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
	# This takes a while, so do it once here and reuse the values as needed
	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
```

@@

	# Replace the actual Tau values with simulated Tau values, using the parameters 
	# estimated for each tree
	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )

@@

	# Replace the actual Tau values with simulated Tau values
	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )

@@

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null simulation, pairwise" ) +
			theme(legend.position="none")

@@

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC simulation, pairwise" ) +
			theme(legend.position="none")

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise). There are several difficulty with interpretting this plot, though. First, it does not just reflect the evolutionary porcesses that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

To better understand this plot (Figure XX_KMRR_Pairwise), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures. First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events. Under the null model, the plot of mean correlation to node age is very similar as for the observed data (Figure XXSim a). As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age. Phylogenetic analysis of the data simulated under the null hypothesis (Figure XXSim b) do not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`), as expected.

We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure XXSim c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`). The phylogenetic analysis of the ortholog conjecture simulation (Figure XXSim c) does reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).

These simulations have several implcations. The pairwise comparisons used by KMRR cannot distinguish betweeh the null hypothesis and ortholog conjecture, as the expected results are the same under both hypotheses. They also serve as a validation of phylogenetic methods as applied to this problem. Our analysis of independent contrasts does not reject the null hypothesis when data are simulated under the null model, and does reject the null hypothesis when the data are simulated under the ortholog conjecture. In contrast to pairwise methods, the phylogenetic analyses make testable predictions based on explicit hyptheses about evolutionary process.

@@

There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi; @Altenhoff:2012ea]. While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk].

Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. The ortholog conjecture does not have to be an all or nothing question, though. It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi]. We just don't find evidence for it when summarizing gene expression across tissues with Tau in these organisms. This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation. These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.

Lack of support for the ortholog conjecture has important biological implications. It indicates that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed. It also has technical implications. Having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].
